Language models as probabilistic cognitive models

### 2023-11-18 03-paper/paper-LLM-probmod.tex
revisions §6

\begin{itemize}
  \item mention multi-set averages
  \begin{itemize}
    \item Let \(\set{I_1, \dots I_m}\) be the multi-set of items that occurred in the human experiment for condition $C$.\footnote{
  By using a multi-set, which may contain a single task multiple times, we produce aggregate predictions for exactly the set of item that the participant group saw, which provides the most fitting counterpart to the human data.
}

  \end{itemize}
\end{itemize}

This section explores different ways of deriving a probabilistic likelihood function for the human data from an LLM at the aggregate, condition-level.
Using log-probabilities for each answer option as a basic item-level score, there are at least three conceivable ways of deriving condition-level predictions by averaging over item-level variation (see Figure~\ref{fig:measures-overview}).
In the following, we first describe the different options of deriving LLM-based predictors.
We then compare all probabilistic models based on their adequacy of explaining the human choice data.

@@

Condition-level predictions are a probability distribution over response types, obtained from item-level scores $\text{S}_{ki}$ by averaging over all items belonging to the relevant condition.
To map non-normalized scores $\myvec{s} = \tuple{s_{1}, \dots, s_{l}}$ to probabilities $\myvec{p} = \tuple{p_{1}, \dots, p_{l}}$ with different degrees of optimization, a common choice is the softmax function with optimality parameter $\alpha$:
%
\begin{align*}
 \text{SoftMax}(\myvec{s}, \alpha) = \myvec{p}, \text{where } p_{i} \propto \expo \left (\alpha p_{i} \right )\,.
\end{align*}
%
The softmax function can furthermore be decomposed into a first step of mapping scores to probabilities, and subsequently reweighing probabilities via a power-law transformation \citep[c.f.,][]{FrankeDegen2023:The-softmax-fun}:
%
\begin{align*}
  \text{SoftMax}(\myvec{s}, \alpha) & = \text{Pow}( \text{SoftMax}(\myvec{s}, 1) ; \alpha) \\
  \text{Pow}(\myvec{p} ; \alpha) & = \myvec{q}, \text{where } q_{i} \propto p_{i}^{\alpha}
\end{align*}
%
This means that there are three conceivable scope-sites for aggregating item-level information as shown in Figure~\ref{fig:measures-overview}.
Narrow-scope aggregation first aggregates the item-level scores, and then transposes the averages into (scaled) probabilities:
%
\begin{align*}
  & P_{\text{cond}}^{\text{SCR}}(r_{i} \mid C, \alpha_{c})
    \propto \expo \left [  \alpha_{c} \ \frac{1}{m} \ \sum_{k=1}^{m} \text{S}_{ki}  \right ]
    \tag*{\textcolor{gray}{[average scores (narrow-scope aggregation)]}}
\end{align*}
%
Wide-scope aggregation, first transposes scores into probabilities, scales them and only aggregates over items last.
\begin{align*}
  % & \textbf{wide-scope aggregation} \\
  & P_{\text{cond}}^{\text{PRB}}(r_{i} \mid C, \alpha_{c})
    = \frac{1}{m} \ \sum_{k=1}^{m} P_{\text{item}}^{\text{LLM}}(y_{ki} \mid C, \alpha_{c})
    \tag*{\textcolor{gray}{[average probabilities (wide-scope aggregation)]}}
\end{align*}
%
Intermediate-scope aggregation performs item-level averaging after mapping scores onto probabilities (using softmax with $\alpha=1$), but before scaling (with a power-law transformation with variable $\alpha$):
\begin{align*}
  % & \textbf{intermediate-scope aggregation} \\
  & P_{\text{cond}}^{\text{WTA}}(r_{i} \mid C, \alpha_{c})
    \propto  \left [ \frac{1}{m} \ \sum_{k=1}^{m} P_{\text{item}}^{\text{WTA}}(y_{ki} \mid C) \right ]^{\alpha_{c}}\,,
    \tag*{\textcolor{gray}{[average WTA (intermediate-scope aggregation)]}}
\end{align*}
where $P_{\text{item}}^{\text{WTA}}(y_{ki} \mid C) = \lim_{\alpha} P_{\text{item}}^{\text{LLM}}(y_{ki} \mid C, \alpha)$.
\bigskip

All three condition-level predictors coincide when there is only one item, of course.
For more items, wide-scope and intermediate-scope aggregation coincide when $\alpha = 1$.
In all other cases, the predictors are not guaranteed to be identical.
Conceptually, the main difference is what each measure considers to be the basic item-level unit to aggregate over.
Narrow-scope aggregation considers raw scores, which are \emph{not} probabilistic predictions at the item level.
Wide-scope aggregation considers $\alpha$-optimized probabilities at the item level, which is compatible with a procedure of sampling, via softmax decoding, item-level answer options.
Wide-scope aggregation contains the ``winner-takes-all'' procedure for accuracy scoring in common benchmark tests for $\alpha \rightarrow \infty$.
Finally, like wide-scope aggregation, intermediate-scope aggregation is also compatible with a sampling based picture, but would use pure decoding (without $\alpha$-optimization) at the item level and reserves the scaling of probability distributions to a stage \emph{after} aggregation.

@@

The three condition-level predictions yield three different probabilistic models, each with two pairs of condition-dependent parameters $\alpha_{c}$, and $\epsilon_{c}$.
Using priors and Bayesian posterior estimation as previously described for the RSA (see Section~\ref{sec:model-pred-from}), we obtain samples from the posterior over parameters and samples from the posterior predictive distributions.
The usual Bayesian summary statistics for posteriors over model parameters are shown in Table~\ref{tab:sumStats}.
What is noteworthy about these estimates is that the $\alpha_{c}$ parameters for the wide-scope model have a very large credible range.
This is because, already for values of $\alpha_{c} = 1$, the item-level predictions are virtually ``winner-takes-all'' choices, selecting the option with the maximum score with probability almost 1 for almost all items.
This implies that, by studying the posterior predictive distribution of the wide-scope model, we are virtually testing condition-level predictions from the standard WTA-strategy used for benchmark testing (see Section~\ref{motivation}).

Figure~\ref{fig:refgame-counts} shows the summary statistics (means and 95\% credible intervals) for each model's posterior predictive distribution.
We find that only the theoretical model (RSA) and the intermediate-scope model passes this ``visual posterior predictive check'' for both conditions; the other two models both overpredict the target choice rate and underpredict the competitor choice rate in the interpretation condition.
To corroborate the visual impression, Table~\ref{tab:sumStats} shows sample-based estimates of Bayesian posterior predictive $p$-values, using likelihood of the observed data as a test statistics.
Consequently, the results from Table~\ref{tab:sumStats} suggest that the narrow-scope model fails on the interpretation data, and is at most borderline compatible with the production data; that the wide-scope model is able to reproduce the production data, but fails on the interpretation data; and that only the intermediate-scope model is able to fully recover the data from both conditions.

@@

These results tell us that not all ways of deriving condition-level predictions by averaging over item-level variation are equally good.
Some approaches clearly fail basic checks for statistical goodness-of-fit.
On the positive side, we also find that there is at least one model with predictors based on LLM-measures, namely the intermediate-scope model, which is able to recover the patterns in the data.
In other words, there is a way of deriving predictor values for condition-level forced-choice probabilities from an LLM such that, when fed into a common linking function (here with parameters $\alpha$ for optimization and $\epsilon$ for random error), the human choice probability can be reconstructed faithfully in its entirety.\footnote{
  A potential worry is that the intermediate-scope model is trivial in the sense that it could have predicted \emph{any} data observation.
  Appendix~\ref{sec:range-prior-pred} shows that this is not the case.
}
This implies that using LLM predictors for probabilistic predictions, such as in a neuro-symbolic model, might be possible if embedded in the proper link functions and if item-level variation is averaged out in the right way.

>>>>

% This section explores different ways of deriving a probabilistic likelihood function for the human data from an LLM at the aggregate, condition-level.
% Using log-probabilities for each answer option as a basic item-level score, there are at least three conceivable ways of deriving condition-level predictions by averaging over item-level variation (see Figure~\ref{fig:measures-overview}).
% In the following, we first describe the different options of deriving LLM-based predictors.
% We then compare all probabilistic models based on their adequacy of explaining the human choice data.

@@

While LLMs do not make conditional-level predictions as such, they can be derived from item-level scores $\text{S}_{ki}$ by averaging over all items belonging to the relevant condition.
There are many ways of averaging item-level information.
Figure~\ref{fig:measures-overview} shows three salient approaches, which differ in what the underlying item-level measure is: the raw scores $\text{S}_{ki}$, the item-level probability derived from it (as introduced in Section~\ref{motivation} and used in Section~\ref{sec:item-level-pred}), or the predictions from the winner-takes-all strategy commonly used in benchmark testing (see Section~\ref{motivation}).

% In general, to map a vector non-normalized scores $\myvec{s} = \tuple{s_{1}, \dots, s_{l}}$ to probabilities $\myvec{p} = \tuple{p_{1}, \dots, p_{l}}$ with different degrees of optimization, a common choice is the softmax function with optimality parameter $\alpha$:
% %
% \begin{align*}
%  \text{SoftMax}(\myvec{s}, \alpha) = \myvec{p}, \text{where } p_{i} \propto \expo \left (\alpha p_{i} \right )\,.
% \end{align*}
% %
% The softmax function can furthermore be decomposed into a first step of mapping scores to probabilities, and subsequently reweighing probabilities via a power-law transformation \citep[c.f.,][]{FrankeDegen2023:The-softmax-fun}:
% %
% \begin{align*}
%   \text{SoftMax}(\myvec{s}, \alpha) & = \text{Pow}( \text{SoftMax}(\myvec{s}, 1) ; \alpha) \\
%   \text{Pow}(\myvec{p} ; \alpha) & = \myvec{q}, \text{where } q_{i} \propto p_{i}^{\alpha}
% \end{align*}

The \emph{average-scores predictor} first aggregates the item-level scores, and then transposes the averages into (scaled) probabilities using the usual parameterize softmax function:\footnote{
  Technically, we average over the multi-set \(\set{I_1, \dots I_m}\) of items that occurred in the human experiment for condition $C$.
  By using a multi-set, which may contain a single task multiple times, we produce aggregate predictions for exactly the set of item that the participant group saw, which provides the most fitting counterpart to the human data.}
%
\begin{align*}
  & P_{\text{cond}}^{\text{SCR}}(r_{i} \mid C, \alpha_{c})
    \propto \expo \left [  \alpha_{c} \ \frac{1}{m} \ \sum_{k=1}^{m} \text{S}_{ki}  \right ]
    \tag*{\textcolor{gray}{[average scores (narrow-scope aggregation)]}}
\end{align*}
%
The \emph{average-probabilities predictor} first transposes scores into probabilities with a parameterized softmax function, and only aggregates over items last.
\begin{align*}
  % & \textbf{wide-scope aggregation} \\
  & P_{\text{cond}}^{\text{PRB}}(r_{i} \mid C, \alpha_{c})
    = \frac{1}{m} \ \sum_{k=1}^{m} P_{\text{item}}^{\text{LLM}}(y_{ki} \mid C, \alpha_{c})
    \tag*{\textcolor{gray}{[average probabilities (wide-scope aggregation)]}}
\end{align*}
%
Finally, the \emph{average-WTA predictor} considers the prediction of the WTA strategy as the basic item-level unit to aggregate over.
To add parameterized scaling to this method a power-law transformation with variable $\alpha$ is a natural choice:
\begin{align*}
  % & \textbf{intermediate-scope aggregation} \\
  & P_{\text{cond}}^{\text{WTA}}(r_{i} \mid C, \alpha_{c})
    \propto  \left [ \frac{1}{m} \ \sum_{k=1}^{m} P_{\text{item}}^{\text{WTA}}(y_{ki} \mid C) \right ]^{\alpha_{c}}\,,
    \tag*{\textcolor{gray}{[average WTA (intermediate-scope aggregation)]}}
\end{align*}
where $P_{\text{item}}^{\text{WTA}}(y_{ki} \mid C) = \lim_{\alpha} P_{\text{item}}^{\text{LLM}}(y_{ki} \mid C, \alpha)$.
\bigskip

The \emph{average-scores} and \emph{average-probabilities} predictions are equivalent if there is only one item, of course.
The prediction of the \emph{average-WTA} method is then the special case of $\alpha \rightarrow \infty$.
For cases with more than one item, the predictions of these different predictors are not guaranteed to be the same.
Conceptually, the \emph{average-probabilities} and the \emph{average-WTA} predictors, but not the \emph{averages-scores} predictor, are compatible with a picture in which condition-level predictions result from the actual predictions at the item level.
The difference between the \emph{average-probabilities} and the \emph{average-WTA} predictors is that the latter is an aggregate of an item-level predictor that is demonstrably bad, as shown in the previous Section~\ref{sec:item-level-pred}.

@@

Using the same approach as for the RSA model in Section~\ref{sec:model-pred-from}, the three condition-level predictions yield three different likelihood functions for the condition-level data.
Using priors and Bayesian posterior estimation as previously described in Section~\ref{sec:model-pred-from}), we obtain samples from the posterior over parameters and samples from the posterior predictive distributions.
The usual Bayesian summary statistics for posteriors over model parameters are shown in Table~\ref{tab:sumStats}.
What is noteworthy is that for the \emph{average-WTA} model, the estimates of $\alpha_{c}$ in the production condition do not rule out, in fact lie close to, the special value $\alpha_{c}=1$, for which the power-law transformation is the identity function.
This means that just averaging WTA-responses at the item-level yields a reasonable predictor for the production data at the condition-level.
However, for the interpretation condition, the value $\alpha_{c}=1$ is clearly outside the range of credible parameter values, so that a simple recipe like ``always average WTA-repsonses without transformation'' is not a viable strategy for good condition-level predictions in general.
It is also worth noting that for the \emph{average-probability} model values of $\alpha_{c}$ close to one are virtually indistinguishable from the item-level predictions of the WTA strategy (see Figure~\ref{fig:closeness-target-item-level}).
This suggests that, for the production data, the aggregating item-level predictions from the WTA-strategy gives good condition-level predictions.
Since the production data is also very biased towards almost exclusively target choices by human participants, this is not too surprising.

Figure~\ref{fig:refgame-counts} shows the summary statistics (means and 95\% credible intervals) for each model's posterior predictive distribution.
We find that only the theoretical model (RSA) and the \emph{average-WTA} model passes this ``visual posterior predictive check'' for both conditions; the other two models both overpredict the target choice rate and underpredict the competitor choice rate in the interpretation condition.
To corroborate the visual impression, Table~\ref{tab:sumStats} shows sample-based estimates of Bayesian posterior predictive $p$-values, using likelihood of the observed data as a test statistics.
Consequently, the results from Table~\ref{tab:sumStats} suggest that the \emph{average-scores} model fails on the interpretation data, and is at most borderline compatible with the production data; that the \emph{average-probabilities} model is able to reproduce the production data, but fails on the interpretation data; and that only the \emph{average-WTA} model is able to fully recover the data from both conditions.

@@

These results tell us that not all ways of deriving condition-level predictions by averaging over item-level variation are equally good.
Some approaches clearly fail basic checks for statistical goodness-of-fit.
On the positive side, we also find that there is at least one model with predictors based on LLM-measures, namely the \emph{average-WTA} model, which is able to recover the patterns in the data.
In other words, there is a way of deriving predictor values for condition-level forced-choice probabilities from an LLM such that, when fed into a common linking function (here with parameters $\alpha$ for optimization and $\epsilon$ for random error), the human choice probability can be reconstructed faithfully in its entirety.\footnote{
  A potential worry is that the \emph{average-WTA} model is trivial in the sense that it could have predicted \emph{any} data observation.
  Appendix~\ref{sec:range-prior-pred} shows that this is not the case.
  \mf{todo: redo these calculations with actual WTA!}
}
This implies that using LLM predictors for probabilistic predictions, such as in a neuro-symbolic model, might be possible if embedded in the proper link functions and if item-level variation is averaged out in the right way.
On the other hand, there is a stark contrast between the results of item-level and condition-level data.
The model that was able to properly fit the condition-level data relies on a model of item-level choice that is clearly incompatible with the human item-level data.
The model seems to be ``right for the wrong reasons.''
